Møil: A web-based administration tool for mail servers backed by relational databases

### 2014-07-29 doc/study/study.tex
Also analyze PostVis Admin.
Finished criteria, started comparison (overview almost finished).

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[ampersand]{easylist}
\usepackage[printonlyused]{acronym}

\usepackage{url}
%\urldef{\urlTcpdump}{\url}{http://www.tcpdump.org/}
\urldef{\urlARSupport}{\url}{http://api.rubyonrails.org/classes/ActiveRecord/Migration.html#class-ActiveRecord::Migration-label-Database+support}

@@

	\section*{Abstract}
		This document compares existing administration user interface solutions
		for mail servers backed by relational databases and introduces yet
		another one, which is called \emph{Møil} and is written in Ruby
		utilizing the \ac{Rails}\footnote{\url{http://rubyonrails.org}}
		framework. It brings the handy possibility of managing the database
		schema with migrations, a lot of beautiful, responsive \acs{CRUD} and
		some other nice features (see page \pageref{sec:moeil:features}). The
		code is available from
		GitHub\footnote{\url{https://github.com/nning/moeil}} licensed under
		AGPLv3 \cite{agpl}.

@@

		Although an extensive evaluation of the existing administration user
		interfaces did not happen before implementing a custom solution, some
		of the programs were still sighted and some of the following criteria
		was still considered. This more systematical and detailed analysis
		shall gain an overview and provide inspiration for the development of
		\emph{Møil}.

		\subsection*{Criteria}
			Not only for the development of \emph{Møil}, some of the criteria
			arose from shortcomings of \emph{postfix.admin} (see page
			\pageref{sec:contestants:postfix.admin}), which is a quite widely
			deployed solution. The criteria will be classified into five
			categories (Data model, features, security \& robustness, user
			interface and miscellaneous) and numbered for easier reference.

@@

			\subsubsection*{Security \& robustness (3)}

			\subsubsection*{User interface (4)}

			\subsubsection*{Miscellaneous (5)}

		\subsection*{The contestants}
			\subsubsection*{modoboa}
				\infotable{http://modoboa.org}{Python}

@@

			\label{sec:contestants:postfix.admin}
			\subsubsection*{postfix.admin}
				\infotable{http://postfixadmin.sourceforge.net}{PHP}

@@

			\subsubsection*{ratuus}
				\infotable{http://www.ratuus.org}{PHP}

			\subsubsection*{VBox.Adm}
				\infotable{http://www.vboxadm.net}{Perl}

			\subsubsection*{ViMbAdmin}
				\infotable{http://www.vimbadmin.net}{PHP}

	\section*{Møil}
		The development of \emph{Møil} has been begun in May, 2013 as an Open
		Source project licensed under the terms of the AGPL, Version 3
		\cite{agpl}.

@@

		\label{sec:moeil:features}
		\subsection*{Features}
			At the time of writing of this document, \emph{Møil} comes with the
			below-mentioned features.

@@

	\newpage
	\section*{Numbered index of criteria}
		\begin{easylist}
			& Data model
			&& Normalized schema
			&& \emph{postfix.admin} compatible
			&& \emph{postfix.admin} import
			&& Updatable (Migrations)
			&& Database independence

@@

			& Security \& robustness
			&& Password hashing
			&& Tests

			& User interface
			&& Responsiveness
			&&&	\ac{CLI} support
			&&& Mobile devices
			&& Usability

			& Miscellaneous
			&& Scalability
			&& Programming language
			&& Containability (e.g. independence of Message Store file system)
		\end{easylist}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[ampersand]{easylist}
\usepackage[printonlyused]{acronym}
\usepackage{lscape}
\usepackage{amssymb}

\usepackage{url}
%\urldef{\urlTcpdump}{\url}{http://www.tcpdump.org/}
\urldef{\urlARSupport}{\url}{http://api.rubyonrails.org/classes/ActiveRecord/Migration.html#class-ActiveRecord::Migration-label-Database+support}
\urldef{\urlPHPBashing}{\url}{http://eev.ee/blog/2012/04/09/php-a-fractal-of-bad-design}

@@

	\section*{Abstract}
		This document compares existing administration user interface solutions
		for mail servers backed by relational databases and introduces yet
		another one, which is called \emph{Møil} and is written in Ruby
		utilizing the \ac{Rails}\footnote{\url{http://rubyonrails.org}}
		framework. It brings the handy possibility of managing the database
		schema with migrations, a lot of beautiful, responsive \acs{CRUD} and
		some other nice features (see \emph{Møil} \textref{sec:moeil:features}).
		The code is available from
		GitHub\footnote{\url{https://github.com/nning/moeil}} licensed under
		AGPLv3 \cite{agpl}.

@@

		Although an extensive evaluation of the existing administration user
		interfaces did not happen before implementing a custom solution, some
		of the programs were still sighted and some of the following criteria
		was still considered. This more systematical and detailed analysis
		shall gain an overview and provide inspiration for the development of
		\emph{Møil}. An in-depth analysis of the source code can not be
		provided here. Instead, the solutions will be compared by a composed set
		of criteria. The current state of the source code was checked out on
		29\textsuperscript{th} July, 2014 from the particular version control
		systems and used for evaluation.

		\subsection*{Criteria}
			\label{sec:evaluation:criteria}
			Not only for the development of \emph{Møil}, some of the criteria
			arose from shortcomings of \emph{postfix.admin} (see page
			\pageref{sec:contestants:postfix.admin}), which is a quite widely
			deployed solution. The criteria will be classified into five
			categories (Data model, features, security \& robustness, user
			interface and miscellaneous) and numbered for easier reference.

			There is some meta-data like website, programming language, last
			update in VCS and number of commits, which is not numbered and
			compared. The used programming language would be a neutral
			attribute but gets a criterium, because before the development of
			\emph{Møil}, a decision was made against deploying further PHP
			applications to accomplish a more maintainable and secure web
			hosting environment\footnote{\urlPHPBashing}.

@@

			\subsubsection*{Security \& robustness (3)}
				For now, only three criteria in this category shall be
				considered, because further security analysis would require an
				in-depth source code review. The \crit{password hashing}{3.1}
				availability and algorithm strength is important against
				unauthorized database access through potential security holes
				in the application. When the database is maliciously read, the
				value of the passwords to the attacker is inversely
				proportional to the time it costs, to crack the password
				hashing. \crit{Test coverage}{3.2} is important for this
				evaluation and for security purposes, because it lowers the
				probability of bugs per lines of code by trend. As
				\crit{Containability}{3.3}, the modular independence (e.g.
				independence of the web interface from the Message Store file
				system) from a security perspective is perceived.

			\subsubsection*{Responsive design (UI) (4)}
				In the days of rapidly increased mobile device usage, a
				responsive design (e.g. different layout for different screen
				sizes) is an important attribute of user interfaces. Usually,
				responsive design means \crit{adapting to mobile devices}{4.1}
				as phones and tablet computers. In this document, also the
				\crit{ability of usage from a \ac{CLI}}{4.2} is considered,
				because it is widespread among system administrators.

		\subsection*{The contestants}
			For introduction of "the contestants", the most important
			attributes and specialities will be listed for each. A more
			detailed comparison of the features can be found under
			\textref{sec:evaluation:overview}.

			\subsubsection*{modoboa}
				\infotable{http://modoboa.org}{Python}{2014-07-28}{1594}

@@

				% Criteria
				% 	2.4.1	No undo.
				%   4.1		Bad mobile experience.

			\subsubsection*{Møil}
				\infotable{https://github.com/nning/moeil}{Ruby}{2014-07-26}{353}

				Further information on \emph{Møil} is to be found under
				\textref{sec:moeil}.

			\subsubsection*{postfix.admin}
			\label{sec:contestants:postfix.admin}
				\infotable{http://postfixadmin.sourceforge.net}{PHP}{2014-07-22}{1682}

@@

				% Criteria
				%	1.4		Migrations one big file upgrade.php with functions
				%			like upgrade_<n>_<db>.
				%	1.5		No systematic database independence (or object
				%			relational mapper at all), only manual support for
				%			mysql and pgsql.
				%	2.4.1	No undo.
				%	3.1		Password hash default md5-crypt.
				%	3.3		Can move mail data if configured and access on Mail
				%			Store, but disabled by default.

			\subsubsection*{PostVis Admin}
				\infotable{http://postvisadmin.sourceforge.net}{PHP}{2010-06-24}{183}
				% SQL injections
				% Badly strutured
				% Code duplication
				% Best example for non-PHP policy
				%
				% Criteria
				%	1.2		Also not normalized like postfix.admin and much
				%			table names and fields equally named.
				%	2.3		Even on a quick view, access control does not seem
				%			effective. Any logged in user (also non-superadmin
				%			users) seem to be able to delete any other user.
				%	3.1		Default md5-crypt.
				%	3.3		Features do not work without shell execution on same
				%			server (e.g. amavisd-release).

			\subsubsection*{ratuus}
				\infotable{http://www.ratuus.org}{PHP}{}{}

				The Development website was not linked correctly; the linked
				hostname did not even exist in DNS, so the URL of the VCS
				could not be determined and the 0.3-r2 version was examined.

				% Bad struture (almost everything inside include/libratuus.php)
				% but way more organized (object oriented, systematic 'routing'
				% and access control) than PostVis Admin, for example. No query
				% sanitization in mailClass->ask_db(). mysql_query() supports
				% multiple queries configuration dependently.
				%
				% Criteria
				%	2.4.1	No undo.
				%	3.1		Only md5-crypt.

			\subsubsection*{VBox.Adm}
				\infotable{http://www.vboxadm.net}{Perl}{2012-12-11}{657}

				% Manual SQL queries, but sanitized. Not easy to get
				% information as for example, if it is database independent.
				%
				% Criteria
				% 	1.5		MySQL dependend in some places (w.g. in maintainance
				%			code), independend in others (e.g. authentication).
				%	3.2		Just custom SMTP proxy for spam maintainance
				%			integration seems to be tested a little.

			\subsubsection*{ViMbAdmin}
				\infotable{http://www.vimbadmin.net}{PHP}{2014-07-23}{337}

				% PHP und ORM!!
				% Criteria
				%	2.4.1	No undo.
				%	2.4.3	Support for access restriction for mailboxes to
				%			SIEVE.
				%	3.1		Default is salted MD5, "more" secure algorithms
				%			only supported with dovecotpw call.
				%	3.3		See 3.1.

			\begin{landscape}
				\subsubsection*{Comparison overview}
				\label{sec:evaluation:overview}
					% 1. modoboa
					% 2. Møil
					% 3. VBox.Adm?
					% 4. postfix.admin
					% 5. ratuus
					% 6. PostVis Admin

					\begin{tabular}{l|cccccc|ccccccccc}
						              & 1 & 1.1 & 1.2  & 1.3 & 1.4   & 1.5  & 2 & 2.1 & 2.2 & 2.3  & 2.4 & 2.4.1 & 2.4.2 & 2.4.3 & 2.4.4 \\
						\hline
						modoboa       &   & \y  &  \n  & \y  &  \y   &  \y  &   & \y  & \y  &  \y  &     & (\y)  & \y    &  \y   & \y    \\
						Møil          &   & \y  &  \n  & \y  &  \y   &  \y  &   & \y  & \y  &  \y  &     &  \y   & \n    &  \n   & \n    \\
						postfix.admin &   & \n  &      &     & (\y)  & (\y) &   & \y  & \y  &  \y  &     & (\y)  & \y    &  \n   & \n    \\
						PostVis Admin &   & \n  & (\y) & \n  &  \n   &  \n  &   & \y  & \n  & (\y) &     &  \n   & \n    &  \n   & \y    \\
						ratuus        &   & \n  &  \y  & \y  &  \n   &  \n  &   & \y  & \y  &  \y  &     & (\y)  & \n    &  \n   & \n    \\
						VBox.Adm      &   & \y  &  \n  & \y  &  \n   & (\y) &   & \y  & \y  &  \y  &     &  \n   & \y    &  \n   & \y    \\
						ViMbAdmin     &   & \y  &  \n  & \n  &       &  \y  &   & \y  & \y  &  \y  &     & (\y)  & \n    & (\n)  & \n    \\
						\hline
					\end{tabular}
					\s

					\noindent
					\begin{tabular}{l|cccc|ccc|ccc}
						              & 3 & 3.1  & 3.2  & 3.3  & 4 & 4.1  & 4.2 \\
						\hline
						modoboa       &   &  \y  &  \y  &  \y  &   & (\y) & \y  \\
						Møil          &   &  \y  &  \y  &  \y  &   &  \y  & \y  \\
						postfix.admin &   & (\y) &  \n  & (\y) &   &  \n  & \y  \\
						PostVis Admin &   & (\y) &  \n  &  \n  &   &  \n  & \n  \\
						ratuus        &   & (\y) &  \n  &  \y  &   &  \n  & \n  \\
						VBox.Adm      &   &  \y  & (\y) &  \y  &   &  \n  & \y  \\
						ViMbAdmin     &   & (\y) &      & (\n) &   &      &     \\
						\hline
					\end{tabular}
					\s

					(See \textref{sec:evaluation:criteria} and
					\textref{sec:appendix:criteria}.)
			\end{landscape}

	\section*{Møil}
	\label{sec:moeil}
		The development of \emph{Møil} has been begun in May, 2013 as an Open
		Source project licensed under the terms of the AGPL, Version 3
		\cite{agpl}.

@@

		\subsection*{Features}
		\label{sec:moeil:features}
			At the time of writing of this document, \emph{Møil} comes with the
			below-mentioned features.

@@

	\newpage
	\section*{Numbered index of criteria}
	\label{sec:appendix:criteria}
		\begin{easylist}
			& Data model
			&& Normalized schema
			&& \emph{postfix.admin} compatible
			&& \emph{postfix.admin} import
			&& Updatable (Migrations)
			&& Database independence

@@

			& Security \& robustness
			&& Password hashing
			&& Test coverage
			&& Containability (e.g. independence of Message Store file system)

			& Responsive design
			&& Mobile devices
			&& \ac{CLI} support
		\end{easylist}
